Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-11-04 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Adding admissible rules in section 2
Reference layout corrected

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \citep{hajek}). Nevertheless, some authors of variants of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of constant domains is motivated by simplicity \citep[footnotes 11 and 14]{and90}. As is well-known, varying domain quantifiers (also known as \emph{actualistic} quantifiers) can be simulated by constant domain quantifiers (also known as \emph{possibilistic} quantifiers) guarded by a primitive \emph{existence} predicate\footnote{Formally, $\forall^e x. F[x] \equiv \forall x. E(x) \imp F[x]$ and $\exists^e x. F[x] \equiv \exists x. E(x) \wedge F[x]$.}; therefore, the formalization of ontological arguments with varying domain quantification would be feasible in the natural deduction calculus presented here as well.

@@

From an axiomatic point of view, modal logics stronger than {\bf K} differ from each other with respect to which axiom schemas they admit in addition to K (and necessitation). Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accessibility relation in Kripke models (See for example \citep{fitting2} or \citep[Section 3.1]{blackburn} on Frame definability): T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, B is satisfied; and in the models of logic {\bf B}, T and B are satisfied.

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{$2$}
\UIC{$\pos \nec A $} \doubleLine
\UIC{$\neg\nec \neg \nec A$}
\AXC{$ $}\RightLabel{$ 1$}
\UIC{$\neg A$}
\AXC{Axiom B}\dashedLine
\UIC{$\neg A\imp \nec\pos\neg A$} \doubleLine
\UIC{$\neg A\imp \nec\neg\nec\neg\neg A$}\RightLabel{$\imp_E$}
\BIC{$\nec\neg\nec\neg\neg A$}\doubleLine
\UIC{$\nec\neg\nec A$} \RightLabel{$\imp_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^1$}
\UIC{$\neg\neg A$} \RightLabel{$\neg\neg_E$}
\UIC{$A$} \RightLabel{$\imp_I^2$}
\UIC{$\pos \nec A\imp A$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
	\AXC{$ $} \RightLabel{1}
	\UIC{$\omega: \rho(\beta)$}
			\AXC{$ $} \RightLabel{3}
			\UIC{$\nec \all x. \neg \rho(x)$} \RightLabel{$\nec_E$}
			\UIC{$\omega: \all x. \neg \rho(x) $} \RightLabel{$\all_E$}
			\UIC{$\omega: \neg \rho(\beta)$} \RightLabel{$\imp_E$}
		\BIC{$\omega: \bot$} \RightLabel{$\bot_E$}
		\UIC{$\omega: \neg (\beta = \beta)$} \dottedLine
		\UIC{$\omega: \beta \neq \beta$} \RightLabel{$\imp_I^1$}
		\UIC{$\omega: \rho(\beta) \imp \beta \neq \beta$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. (\rho(x) \imp x \neq x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec \all x. (\rho(x) \imp x \neq x)\scott$}  \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x \neq x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x \neq x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x \neq x) \imp P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
		\BIC{$[\subproof_2] \qquad P(\lambda x. x \neq x)\scott$} 
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
		\AXC{$\subproof_2$} \dashedLine
		\UIC{$P(\lambda x. x \neq x)\scott$}    
    		\AXC{$\subproof_1$} \dashedLine
			  \UIC{$P(\lambda x. x = x)\scott$} 
					\AXC{Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
			\BIC{$[\subproof_3] \qquad \bot$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec \neg \ex x. \rho(x)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x. \rho(x)$}
        \AXC{$ $} \RightLabel{2}
        \UIC{$\omega: \rho(\alpha)$} \RightLabel{$\ex_I$}
		\UIC{$\omega: \ex x. \rho(x)$} \RightLabel{$\imp_E$}
    \BIC{$\omega: \bot$} \RightLabel{$\imp_I^2$}
	\UIC{$\omega: \neg \rho(\alpha)$} \RightLabel{$\all_I$}
	\UIC{$\omega: \all x. \neg \rho(x)$} \RightLabel{$\nec_I$}
	\UIC{$\nec \all x. \neg \rho(x) \scott$}
	        \AXC{$\subproof_3 $} \dashedLine
	        \UIC{$\bot$} \RightLabel{$\imp_I^3$}
    		\UIC{$\neg \nec \all x. \neg \rho(x)$} \RightLabel{$\imp_E$}
  		\BIC{$\bot$} \RightLabel{$\imp_I^4$}
  		\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
  		\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^5$}
  		\UIC{$ P(\rho) \imp \pos \ex x.\rho(x)$} \RightLabel{$\all_I$}
      \UIC{$ \all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x)) $}
\end{prooftree}
\end{small}
\end{proof}

@@

		\AXC{$ $} \RightLabel{1}
		\UIC{$\neg P(\rho)$} 
				\AXC{Axiom 1} \dashedLine
        \UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
        \UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\wedge_E$}
        \UIC{$\neg P(\rho) \imp P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
			\BIC{$P(\lambda x. \neg \rho(x))$}

					\AXC{$ $} \RightLabel{5}
					\UIC{$G(\gamma)$} \dottedLine
					\UIC{$\all \varphi. P(\varphi) \imp \varphi(\gamma)$} \RightLabel{$\all_E$}
					\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$} \RightLabel{$\imp_E$}
				\BIC{$\neg \rho(\gamma)$} \RightLabel{$\imp_E$}      
	   \BIC{$\bot$} \RightLabel{$\imp_I^1$}
	   \UIC{$\neg\neg P(\rho)$} \RightLabel{$\neg\neg_E$}
	   \UIC{$[\subproof_4] \qquad P(\rho)$}
	   \end{prooftree}
\end{small}

@@

\AXC{$ $} \RightLabel{1}
\UIC{$\pos \ex x. G(x)$} 
		\AXC{Lemma 1} \dashedLine
		\UIC{$\omega: \ex x. G(x) \imp \nec \ex x. G(x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\ex x. G(x) \imp \nec \ex x. G(x)) $}
				\AXC{Distribution Principle}
			\BIC{$\pos \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
	\BIC{$\pos \nec \ex x. G(x)\scott$}
			\AXC{S5 Iteration Principle} \dashedLine
			\UIC{$\pos \nec \ex x. G(x) \imp \nec \ex x. G(x)$}
		\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
		\UIC{$\pos \ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{scriptsize}

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$P(\rho)$} 
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.\rho(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.\rho(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: \rho(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.\rho(x) $} \RightLabel{$\imp_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg \rho(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  \rho(x) \imp \neg \rho(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(\rho(x) \imp \neg \rho(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(\rho(x) \imp \neg \rho(x)) $}\RightLabel{$\wedge_I$}
\BIC{$ P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))$}
\AXC{Axiom \ref{A2} for $\rho$ and $\lambda x. \neg \rho(x)$} \dashedLine
\UIC{$(P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))) \imp P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\BIC{$[\subproof_5]\qquad P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\end{prooftree}
\end{scriptsize}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_5$} \dashedLine
\UIC{$P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\wedge_E$}
\UIC{$ P(\lambda x. \neg \rho(x)) \imp \neg P(\rho) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(\rho) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.\rho(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x)) $}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\neg P(\rho)$} 
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\wedge_E$}
\UIC{$\neg P(\rho) \imp P(\lambda x. \neg \rho(x))$}\RightLabel{$\imp_E$}
\BIC{$P(\lambda x. \neg \rho(x))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\neg \rho(\gamma)$}
\AXC{$ $} \RightLabel{3}
\UIC{$\rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^2$}
\UIC{$\neg\neg P(\rho)$}\RightLabel{$\neg\neg_E$}
\UIC{$[\subproof_6] \qquad P(\rho)$}
\end{prooftree}
\end{small}

@@

Since the iteration principle requires the strong modal logic {\bf S5} while Brouwer's reduction principle requires only the much weaker modal logic {\bf KB},  the possibility of proving Theorem 3 using Brouwer's reduction principle instead of {\bf S5}'s iteration principle is philosophically profound. There have been many discussions of what modal axioms are needed for the ontological argument. In their manuscripts, G\"odel and Scott do not write explicitly which logic they use. However, a proof step in their notes clearly relies on {\bf S5}'s \emph{iteration principle}.
Anderson \citep[footnote 5]{and90} cites several articles questioning the adequacy of {\bf S5} for the ontological argument and conjectures that the ontological argument could be restricted to the weaker modal logic {\bf B}. Sobel \citep{sobel2}[p. 152] acknowledges Anderson's conjecture, claims that it is correct and tries to informally explain how his formal proofs could be modified to rely on {\bf B} instead of {\bf S5}. Nevertheless, neither Anderson nor Sobel presented any formal proof relying only on {\bf B}. The new formal proof presented in this section relies on {\bf KB}, which is weaker than {\bf B}.

@@

\begin{small}
\begin{prooftree}
\AXC{$ \subproof_8 $}\dashedLine
\UIC{$\ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\ex y.G(y)$} \RightLabel{$\imp_E$}
\BIC{$ A \imp \nec \all x. (G(x) \imp A)$} \doubleLine
\UIC{$ A \imp \nec (\ex x. G(x) \imp A)$}
\AXC{$ $} \RightLabel{1}
\UIC{$A$} \RightLabel{$\imp_E$}
\BIC{$ \nec (\ex x. G(x) \imp A)$} \RightLabel{$\nec_E$}
\UIC{$ \omega:  \ex x. G(x) \imp A$} \RightLabel{$\imp_E$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\omega: A$} \RightLabel{$\nec_I$}
\UIC{$ \nec A$} \RightLabel{$\imp_I^1$}
   \UIC{$A\imp \nec A$}
\end{prooftree}
\end{small}

In G\"odel's ontological proof, we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.
Fitting, for instance, has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep{fitting}[Sections 11.9 and 11.10]. Koons \citep{koons}, Anderson \citep{and90}[p. 292] and Sobel \citep{sobel2}[p. 133] also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate. Two alternative and more conservative solutions, requiring only modifications of the axioms and definitions but no change in the logic, are discussed in greater detail in the next sections.

>>>>

\begin{calculus}
{Admissible propositional rules}
{fig:AdmissiblePropositionalRules}

\s\s
\infer[\neg_E]{\bot}{ \neg A & A }
\s\s
\infer[\biimp_E]{A}{A\biimp B & B  }
\s\s
\infer[\biimp_E]{B}{A\biimp B & A  }

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants \citetext{e.g. \citealt{hajek}}. Nevertheless, some authors of variants of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of constant domains is motivated by simplicity \citep[footnotes 11 and 14]{and90}. As is well-known, varying domain quantifiers (also known as \emph{actualistic} quantifiers) can be simulated by constant domain quantifiers (also known as \emph{possibilistic} quantifiers) guarded by a primitive \emph{existence} predicate\footnote{Formally, $\forall^e x. F[x] \equiv \forall x. E(x) \imp F[x]$ and $\exists^e x. F[x] \equiv \exists x. E(x) \wedge F[x]$.}; therefore, the formalization of ontological arguments with varying domain quantification would be feasible in the natural deduction calculus presented here as well.

@@

From an axiomatic point of view, modal logics stronger than {\bf K} differ from each other with respect to which axiom schemas they admit in addition to K (and necessitation). Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accessibility relation in Kripke models \citetext{see e.g. \citealt{fitting2} or \citealt[Section 3.1]{blackburn} on Frame definability}: T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, B is satisfied; and in the models of logic {\bf B}, T and B are satisfied.

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{$2$}
\UIC{$\pos \nec A $} \doubleLine
\UIC{$\neg\nec \neg \nec A$}
\AXC{$ $}\RightLabel{$ 1$}
\UIC{$\neg A$}
\AXC{Axiom B}\dashedLine
\UIC{$\neg A\imp \nec\pos\neg A$} \doubleLine
\UIC{$\neg A\imp \nec\neg\nec\neg\neg A$}\RightLabel{$\imp_E$}
\BIC{$\nec\neg\nec\neg\neg A$}\doubleLine
\UIC{$\nec\neg\nec A$} \RightLabel{$\neg_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^1$}
\UIC{$\neg\neg A$} \RightLabel{$\neg\neg_E$}
\UIC{$A$} \RightLabel{$\imp_I^2$}
\UIC{$\pos \nec A\imp A$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
	\AXC{$ $} \RightLabel{1}
	\UIC{$\omega: \rho(\beta)$}
			\AXC{$ $} \RightLabel{3}
			\UIC{$\nec \all x. \neg \rho(x)$} \RightLabel{$\nec_E$}
			\UIC{$\omega: \all x. \neg \rho(x) $} \RightLabel{$\all_E$}
			\UIC{$\omega: \neg \rho(\beta)$} \RightLabel{$\neg_E$}
		\BIC{$\omega: \bot$} \RightLabel{$\bot_E$}
		\UIC{$\omega: \neg (\beta = \beta)$} \dottedLine
		\UIC{$\omega: \beta \neq \beta$} \RightLabel{$\imp_I^1$}
		\UIC{$\omega: \rho(\beta) \imp \beta \neq \beta$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. (\rho(x) \imp x \neq x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec \all x. (\rho(x) \imp x \neq x)\scott$}  \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x \neq x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x \neq x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x \neq x) \imp P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
		\BIC{$[\subproof_2] \qquad P(\lambda x. x \neq x)\scott$} 
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
		\AXC{$\subproof_2$} \dashedLine
		\UIC{$P(\lambda x. x \neq x)\scott$}    
    		\AXC{$\subproof_1$} \dashedLine
			  \UIC{$P(\lambda x. x = x)\scott$} 
					\AXC{Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\neg_E$}
			\BIC{$[\subproof_3] \qquad \bot$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec \neg \ex x. \rho(x)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x. \rho(x)$}
        \AXC{$ $} \RightLabel{2}
        \UIC{$\omega: \rho(\alpha)$} \RightLabel{$\ex_I$}
		\UIC{$\omega: \ex x. \rho(x)$} \RightLabel{$\neg_E$}
    \BIC{$\omega: \bot$} \RightLabel{$\imp_I^2$}
	\UIC{$\omega: \neg \rho(\alpha)$} \RightLabel{$\all_I$}
	\UIC{$\omega: \all x. \neg \rho(x)$} \RightLabel{$\nec_I$}
	\UIC{$\nec \all x. \neg \rho(x) \scott$}
	        \AXC{$\subproof_3 $} \dashedLine
	        \UIC{$\bot$} \RightLabel{$\imp_I^3$}
    		\UIC{$\neg \nec \all x. \neg \rho(x)$} \RightLabel{$\neg_E$}
  		\BIC{$\bot$} \RightLabel{$\imp_I^4$}
  		\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
  		\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^5$}
  		\UIC{$ P(\rho) \imp \pos \ex x.\rho(x)$} \RightLabel{$\all_I$}
      \UIC{$ \all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x)) $}
\end{prooftree}
\end{small}
\end{proof}

@@

		\AXC{$ $} \RightLabel{1}
		\UIC{$\neg P(\rho)$} 
				\AXC{Axiom 1} \dashedLine
        \UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
        \UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
			\BIC{$P(\lambda x. \neg \rho(x))$}

					\AXC{$ $} \RightLabel{5}
					\UIC{$G(\gamma)$} \dottedLine
					\UIC{$\all \varphi. P(\varphi) \imp \varphi(\gamma)$} \RightLabel{$\all_E$}
					\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$} \RightLabel{$\imp_E$}
				\BIC{$\neg \rho(\gamma)$} \RightLabel{$\neg_E$}      
	   \BIC{$\bot$} \RightLabel{$\imp_I^1$}
	   \UIC{$\neg\neg P(\rho)$} \RightLabel{$\neg\neg_E$}
	   \UIC{$[\subproof_4] \qquad P(\rho)$}
	   \end{prooftree}
\end{small}

@@

\AXC{$ $} \RightLabel{1}
\UIC{$\pos \ex x. G(x)$} 
		\AXC{Lemma 1} \dashedLine
		\UIC{$\omega: \ex x. G(x) \imp \nec \ex x. G(x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\ex x. G(x) \imp \nec \ex x. G(x)) $}
				\AXC{Distribution Principle}
			\BIC{$\pos \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
	\BIC{$\pos \nec \ex x. G(x)\scott$}
			\AXC{S5 Iteration Principle} \dashedLine
			\UIC{$\pos \nec \ex x. G(x) \imp \nec \ex x. G(x)$}\RightLabel{$\imp_E$}
		\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
		\UIC{$\pos \ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{scriptsize}

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$P(\rho)$} 
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.\rho(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.\rho(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: \rho(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.\rho(x) $} \RightLabel{$\neg_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg \rho(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  \rho(x) \imp \neg \rho(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(\rho(x) \imp \neg \rho(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(\rho(x) \imp \neg \rho(x)) $}\RightLabel{$\wedge_I$}
\BIC{$ P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))$}
\AXC{Axiom \ref{A2} for $\rho$ and $\lambda x. \neg \rho(x)$} \dashedLine
\UIC{$(P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))) \imp P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\BIC{$[\subproof_5]\qquad P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\end{prooftree}
\end{scriptsize}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_5$} \dashedLine
\UIC{$P(\lambda x. \neg \rho(x))$} %\RightLabel{$\imp_E$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
\BIC{$\neg P(\rho) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(\rho)$} \RightLabel{$\neg_E$}
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.\rho(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x)) $}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\neg P(\rho)$} 
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
\BIC{$P(\lambda x. \neg \rho(x))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\neg \rho(\gamma)$}
\AXC{$ $} \RightLabel{3}
\UIC{$\rho(\gamma)$} \RightLabel{$\neg_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^2$}
\UIC{$\neg\neg P(\rho)$}\RightLabel{$\neg\neg_E$}
\UIC{$[\subproof_6] \qquad P(\rho)$}
\end{prooftree}
\end{small}

@@

Since the iteration principle requires the strong modal logic {\bf S5} while Brouwer's reduction principle requires only the much weaker modal logic {\bf KB},  the possibility of proving Theorem 3 using Brouwer's reduction principle instead of {\bf S5}'s iteration principle is philosophically profound. There have been many discussions of what modal axioms are needed for the ontological argument. In their manuscripts, G\"odel and Scott do not write explicitly which logic they use. However, a proof step in their notes clearly relies on {\bf S5}'s \emph{iteration principle}.
Anderson \citep[footnote 5]{and90} cites several articles questioning the adequacy of {\bf S5} for the ontological argument and conjectures that the ontological argument could be restricted to the weaker modal logic {\bf B}. Sobel \citep[p. 152]{sobel2} acknowledges Anderson's conjecture, claims that it is correct and tries to informally explain how his formal proofs could be modified to rely on {\bf B} instead of {\bf S5}. Nevertheless, neither Anderson nor Sobel presented any formal proof relying only on {\bf B}. The new formal proof presented in this section relies on {\bf KB}, which is weaker than {\bf B}.

@@

\begin{small}
\begin{prooftree}
\AXC{$ \subproof_8 $}\dashedLine
\UIC{$\ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\ex y.G(y)$} \RightLabel{$\imp_E$}
\BIC{$ A \imp \nec \all x. (G(x) \imp A)$} \doubleLine
\UIC{$ A \imp \nec (\ex x. G(x) \imp A)$}
\AXC{$ $} \RightLabel{1}
\UIC{$A$} \RightLabel{$\imp_E$}
\BIC{$ \nec (\ex x. G(x) \imp A)$} \RightLabel{$\nec_E$}
\UIC{$ \omega:  \ex x. G(x) \imp A$} %\RightLabel{$\imp_E$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\omega: A$} \RightLabel{$\nec_I$}
\UIC{$ \nec A$} \RightLabel{$\imp_I^1$}
   \UIC{$A\imp \nec A$}
\end{prooftree}
\end{small}

In G\"odel's ontological proof, we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.
Fitting, for instance, has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep[Sections 11.9 and 11.10]{fitting}. Koons \citep{koons}, Anderson \citep[p. 292]{and90} and Sobel \citep[p. 133]{sobel2} also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate. Two alternative and more conservative solutions, requiring only modifications of the axioms and definitions but no change in the logic, are discussed in greater detail in the next sections.
